### 2010-09-30 Thesis/Chapter3.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Solution interpolation}	\label{chap3:solutionInterpolation}
As we have seen, the finite element method is based on finding an approximate solution over each simple element rather than the whole domain. Any continuous function $ f $ may be approximated by a linear combination of known functions $ \phi_i $ and undetermined coefficients $ c_i $:
\begin{equation}
f \approx \tilde{f} = \sum_{i=1}^n c_i \phi_i .
\end{equation}
Moreover, the approximation solution $ u^e $within the element $\Omega_e $ must fullfill certain conditions of continuity in order to be convergent to the actual solution $ u $ as the number of elements is increased. The finite element method approximates the solution by the following polynomial expression:
\begin{equation}
\label{chap3:polynom}
u^e(x) = \sum_{i=1}^n u^e_i \psi^e_i(x) 
\end{equation}
where $ \Omega_e $ is a one-dimensional element\footnote{For the sake of simplicity, all derivations are carried out in 1D but remain valid for each component in higher dimensions.}, $ x $ a position within this element, $ u^e_i $ are the values of the solution $ u $ at the nodes and $ \psi^e_i(x) $ the approximation functions over the element. This particular form will be assumed for brevity but the interested reader may refer to \cite{Reddy93} for demonstration. Note that $ u^e_i  $ plays the role of the undetermined coefficients and $ \psi^e_i $ the role of approximation functions. Writing the approximation in terms of the nodal values of the solution is necessitated by the fact that we require the solution to be the same at points common to the elements in order to connect the approximate solution from each element and form a continuous solution over the whole domain. The approximation $ \psi^e_i $ can be linear or higher-order polynomial expressions and are called interpolation functions. 
%They are also commonly named \emph{shape functions} because they define the `shape' that a variable can take within an element (linear, quadratic etc.). 
Depending on the degree of polynomial approximation used to represent the solution, additional nodes may be identified inside the element. It is worth noting that the type of interpolation directly affects the accuracy of the solution. In other words, finding the solution to the problem consists of merely figuring out the values of the sought variable at every node of the mesh. Its value at any other point of the domain may then be deducted by interpolation within elements.

@@

	\subsection{Natural coordinates}
The natural coordinate system allows us to map every element into a typical and simpler element. As an example, an 8-node hexahedron element is shown in \fig{chap3:fig-naturalCoordinatesHexa} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems. While in its natural coordinate system the element is a regular aligned cube, in the global coordinate system the element may assume any admissible arbitrary form. Essentially, admissible means that the element must not be too distorted, and must certainly not fold back on itself. Another example is given \fig{chap3:fig-naturalCoordinatesTetra} with a tetrahedral element. The components of each node has a simple expression in natural coordinates, usually $0, 1$ or $-1$ and the centre of the coordinate system is taken at one of the nodes or at the centre of the element. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter3/naturalCoordinatesHexa.pdf}
\end{center}
\caption[Natural coordinates of a hexaedron]{Hexahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
\label{chap3:fig-naturalCoordinatesHexa}
\end{figure}
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter3/naturalCoordinatesTetra.pdf}
\end{center}
\caption[Natural coordinates of a tetrahedron]{Tetrahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
\label{chap3:fig-naturalCoordinatesTetra}
\end{figure}

@@

The next step is therefore to derive the element equations themselves. Obviously these equations depend on the type of problem that we seek to solve. From now on, we will only consider the context of continuum mechanics since our overall goal is the modelling of organ deformation. However, a similar approach can be used in the other fields of physics. In continuum mechanics, applying Newton's second law yields a partial differential system of equations. Such equations are called \emph{strong forms}. The strong form, in contrast to a \emph{weak form}, requires strong continuity on the dependent field variables, such as displacements \citep{Liu03}. The functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from. Obtaining the exact solution for a strong form is usually very difficult for practical engineering problems. We know that the finite element method can be used to find an approximated solution but the method usually works well only with regular geometry and boundary conditions.

A weak form is often created using energy principles or weighted residual methods. The weak form is often an integral form and requires a weaker continuity on the field variables. Due to the weaker requirement on the field variables, and the integral operation, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry. Hence, the weak form is usually preferred for obtaining an approximate solution. Using the weak form usually leads to a set of well-behaved algebraic system equations, if the problem domain is discretised properly into elements.

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particulary in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Space-time coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

However, only solvers relevant to the context of finite element methods will be discussed. Indeed the matrix $ \mathbf{A} $ takes a specific form in problems solved by finite element methods. The matrix $ \mathbf{A} $ is \emph{sparse}. A sparse matrix is merely a matrix populated mainly with zeros. An example is given by \fig{chap3:fig-sparse} for a finite element problem in two dimensions.
\begin{figure}
\begin{center}
\includegraphics[width=13cm]{chapter3/sparse_matrix.pdf}
\end{center}
\caption[Illustration of a sparse matrix for a 2D finite element problem]{Illustration of a sparse matrix for a two-dimensional finite element problem. Each non-zero elements of the matrix is represented in black. Image courtesy of Oleg Alexandrov.}
\label{chap3:fig-sparse}
\end{figure}
Because of the peculiar structure of the matrix involved in finite element computations, methods have been specifically designed to take advantage of this structure. Indeed, it is wasteful to use general methods of linear algebra on such problems, because most of the arithmetic operations devoted to solving the set of equations or inverting the matrix involve zero operands. Furthermore, it is wasteful to reserve storage for zero elements.

%The choice of a resolution method depends on the problem (linear or non-linear) and the properties of the stiffness matrix.

@@

One of the simplest preconditioner is the \emph{Jacobi preconditioner} where the preconditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of a LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details). 
\OFF

>>>>

	\subsection{Solution interpolation}	\label{chap3:solutionInterpolation}
As we have seen, the finite element method is based on finding an approximate solution over each simple element rather than the whole domain. Any continuous function $ f $ may be approximated by a linear combination of known functions $ \phi_i $ and undetermined coefficients $ c_i $:
\begin{equation}
f \approx \tilde{f} = \sum_{i=1}^n c_i \phi_i .
\end{equation}
Moreover, the approximation solution $ u^e $within the element $\Omega_e $ must fullfill certain conditions of continuity in order to be convergent to the actual solution $ u $ as the number of elements is increased. The finite element method approximates the solution by the following polynomial expression:
\begin{equation}
\label{chap3:polynom}
u^e(x) = \sum_{i=1}^n u^e_i \psi^e_i(x) 
\end{equation}
where $ \Omega_e $ is a one-dimensional element\footnote{For the sake of simplicity, all derivations are carried out in 1D but remain valid for each component in higher dimensions.}, $ x $ a position within this element, $ u^e_i $ are the values of the solution $ u $ at the nodes and $ \psi^e_i(x) $ the approximation functions over the element. This particular form will be assumed for brevity but the interested reader may refer to \cite{Reddy93} for demonstration. Note that $ u^e_i  $ plays the role of the undetermined coefficients and $ \psi^e_i $ the role of approximation functions. Writing the approximation in terms of the nodal values of the solution is necessitated by the fact that we require the solution to be the same at points common to the elements in order to connect the approximate solution from each element and form a continuous solution over the whole domain. The approximation $ \psi^e_i $ can be linear or higher-order polynomial expressions and are called interpolation functions. 
Depending on the degree of polynomial approximation used to represent the solution, additional nodes may be identified inside the element. It is worth noting that the type of interpolation directly affects the accuracy of the solution. In other words, finding the solution to the problem consists of merely figuring out the values of the sought variable at every node of the mesh. Its value at any other point of the domain may then be deducted by interpolation within elements.

@@

	\subsection{Natural coordinates}
The natural coordinate system allows us to map every element into a typical and simpler element. As an example, an 8-node hexahedron element is shown in \fig{chap3:fig-naturalCoordinatesHexa} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems. While in its natural coordinate system the element is a regular aligned cube, in the global coordinate system the element may assume any admissible arbitrary form. Essentially, admissible means that the element must not be too distorted, and must certainly not fold back on itself. Another example is given \fig{chap3:fig-naturalCoordinatesTetra} with a tetrahedral element. The components of each node has a simple expression in natural coordinates, usually $0, 1$ or $-1$ and the centre of the coordinate system is taken at one of the nodes or at the centre of the element. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesHexa1.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesHexa2.pdf}}
\caption[Natural coordinates of a hexaedron]{Hexahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
\label{chap3:fig-naturalCoordinatesHexa}
\end{figure}
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesTetra1.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesTetra2.pdf}}
\caption[Natural coordinates of a tetrahedron]{Tetrahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
\label{chap3:fig-naturalCoordinatesTetra}
\end{figure}

@@

The next step is therefore to derive the element equations themselves. Obviously these equations depend on the type of problem that we seek to solve. From now on, we will only consider the context of continuum mechanics since our overall goal is the modelling of organ deformation. However, a similar approach can be used in the other fields of physics. In continuum mechanics, applying Newton's second law yields a partial differential system of equations. Such equations are called \emph{strong forms}. The strong form, in contrast to a \emph{weak form}, requires strong continuity on the dependent field variables, such as displacements \citep{Liu03}. \ON Indeed, \OFF the functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from. \ON While the finite element method can be used to find an approximated solution for a strong form,  the method usually works well only with regular geometry and boundary conditions.

Consequently, \OFF a weak form is often created using energy principles or weighted residual methods. The weak form is frequently an integral form and requires a weaker continuity on the field variables. \ON Hence, \OFF a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry. \ON For this reason, \OFF the weak form is usually preferred \ON over the strong form for obtaining an approximate solution of a practical engineering problem \OFF.

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particulary in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

However, only solvers relevant to the context of finite element methods will be discussed. Indeed the matrix $ \mathbf{A} $ takes a specific form in problems solved by finite element methods. The matrix $ \mathbf{A} $ is \emph{sparse}. A sparse matrix is merely a matrix populated mainly with zeros. An example is given by \fig{chap3:fig-sparse} for a finite element problem in two dimensions.
\begin{figure}[ht]
\begin{center}
\includegraphics[width=10cm]{chapter3/sparse_matrix.pdf}
\end{center}
\caption[Illustration of a sparse matrix for a 2D finite element problem]{Illustration of a sparse matrix for a two-dimensional finite element problem. Each non-zero elements of the matrix is represented in black. Image courtesy of Oleg Alexandrov.}
\label{chap3:fig-sparse}
\end{figure}
Because of the peculiar structure of the matrix involved in finite element computations, methods have been specifically designed to take advantage of this structure. Indeed, it is wasteful to use general methods of linear algebra on such problems, because most of the arithmetic operations devoted to solving the set of equations or inverting the matrix involve zero operands. Furthermore, it is wasteful to reserve storage for zero elements.

@@

One of the simplest preconditioner is the \emph{Jacobi preconditioner} where the preconditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details). 
\OFF

### 2010-09-30 Thesis/Chapter4.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\ON
The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter \label{chap1}. Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

@@

The use of the finite element method in computer graphics was pioneered by \cite{Terzopoulos87} for simulations of elastic deformations. They derived a geometrically non-linear strain tensor from differential geometry theory. Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner (see \fig{chap4:fig-ball}).
 %
\begin{figure}[h]
\begin{center}
\includegraphics[width=7cm]{chapter4/ball.pdf}
\end{center}
\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}. \TODO{to be deleted?}} 
\label{chap4:fig-ball}
\end{figure}	
In order to speed up the computations, linear strain measures started to be used (see section~\ref{chap2:infinitesimalStrainTensor}). For instance, \cite{Gourret89} make use of a linear finite element method to model the deformation of grasped objects. The equations are however solved in static for simplicity. An approach to solve the system of equations in dynamic in a reasonable amount of time is to decrease the number of elements. This reduction must be compensated by the use of high-order elements, that is the degree of the shape functions used in the interpolation of the solution is higher to maintain the overall accuracy. This approach was applied to an elaborate model of a muscle by \cite{Chen92} and to modelling skin in a simulator of plastic surgery by \cite{Pieper92} where they both use 20-node brick elements. Yet, finite element modelling remained very computationally expensive and the interactivity required for medical simulators was still to be reached.

@@

\cite{Cotin00} proposed a hybrid model allowing real-time cutting. They combined a quasi-static precomputed model, which is very efficient but does not allow topology changes, with a tensor-mass model which requires more computation but authorises cutting. The key idea is to separate a same structure into two parts: one susceptible to be cut during the surgery procedure and one that will not be submitted to any topological change. While the former will be simulated with the tensor-mass model, the deformation of the latter will be computed using the precomputed model. Since both linearly elastic models follow the same physical law, the combination of these two models should behave exactly as a global, linearly elastic model. To achieve this goal, \citeauthor{Cotin00} describe how to impose the additional boundary conditions at the connection nodes for the global model to be consistent in terms of forces and displacements. The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver). They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see \fig{chap4:fig-hybridModel}).
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model. $ 18 $\% of the mesh was modelled with the tensor-mass model (green) and therefore could be cut out. The rest of the mesh was simulated with the precomputed finite element model. Image courtesy of \cite{Cotin00}. \TODO{Stéphane> c'est ptet moche comme image, mais c'est pourtant bien ça dans le papier... Et même dans ton HDR tu utilises cette image... regarde page 46 et 50... Donc ben c'est pas de ma faute si c'est moche ;-) }} 
\label{chap4:fig-hybridModel}
\end{figure}

@@

Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. \TODO{add PhD of Brian Lee... But I don't have it, and is it published yet?}
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
\end{center}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

\TODO{+add paper ISBMS 2010?}

\ON
Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically. They combined a new strain-stress tensor computation scheme with smoothed particule hydrodynamics (SPH) method. The parameters of the tensor relating strain and stress are determined through mechanical experiments.

\OFF

>>>>

\ON
The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter (see chapter \ref{chap1}). Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

@@

The use of the finite element method in computer graphics was pioneered by \cite{Terzopoulos87} for simulations of elastic deformations. They derived a geometrically non-linear strain tensor from differential geometry theory. Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner.
%(see \fig{chap4:fig-ball}).
 %
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=7cm]{chapter4/ball.pdf}
%\end{center}
%\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}. \TODO{to be deleted?}} 
%\label{chap4:fig-ball}
%\end{figure}

In order to speed up the computations, linear strain measures started to be used (see section~\ref{chap2:infinitesimalStrainTensor}). For instance, \cite{Gourret89} make use of a linear finite element method to model the deformation of grasped objects. The equations are however solved in static for simplicity. An approach to solve the system of equations in dynamic in a reasonable amount of time is to decrease the number of elements. This reduction must be compensated by the use of high-order elements, that is the degree of the shape functions used in the interpolation of the solution is higher to maintain the overall accuracy. This approach was applied to an elaborate model of a muscle by \cite{Chen92} and to modelling skin in a simulator of plastic surgery by \cite{Pieper92} where they both use 20-node brick elements. Yet, finite element modelling remained very computationally expensive and the interactivity required for medical simulators was still to be reached.

@@

\cite{Cotin00} proposed a hybrid model allowing real-time cutting. They combined a quasi-static precomputed model, which is very efficient but does not allow topology changes, with a tensor-mass model which requires more computation but authorises cutting. The key idea is to separate a same structure into two parts: one susceptible to be cut during the surgery procedure and one that will not be submitted to any topological change. While the former will be simulated with the tensor-mass model, the deformation of the latter will be computed using the precomputed model. Since both linearly elastic models follow the same physical law, the combination of these two models should behave exactly as a global, linearly elastic model. To achieve this goal, \citeauthor{Cotin00} describe how to impose the additional boundary conditions at the connection nodes for the global model to be consistent in terms of forces and displacements. The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver). They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see \fig{chap4:fig-hybridModel}).
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model. $ 18 $\% of the mesh was modelled with the tensor-mass model (green) and therefore could be cut out. The rest of the mesh was simulated with the precomputed finite element model. Image courtesy of \cite{Cotin00}. \TODO{Stéphane> tu dis que l'image est moche, mais c'est pourtant bien celle présente dans le papier... Et même dans ton HDR tu utilises cette image... regarde page 46 et 50... Mine de rien ça fait plus de 10 ans ;-) }} 
\label{chap4:fig-hybridModel}
\end{figure}

@@

Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. \TODO{add PhD of Brian Lee... Mais je l'ai pas sa thèse... et tu m'as pas dit que tu étais en train de la reviewer ? Du coup c'est pas encore publique, c'est difficile de la citer, non ?}
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
\end{center}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

\ON Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically in a laparoscopic simulator. They combined a new strain-stress tensor computation scheme with smoothed particule hydrodynamics (SPH) method. The parameters of the tensor relating strain and stress were determined through mechanical experiments. \OFF
